### 2025-04-03 scripts/Paper-analysis.Rmd
add Figure 1 into main script

```{r}
library(ggplot2)
library(knitr)
library(xtable)
library(dplyr)
library(tidyr)
```

@@

```{r}
# Pangenome statistics for PTUs
ptu = read.csv('../data/PTU-pangenome-stats.csv')

@@

# Maximum counts
d.reduced.median.plot = d.reduced.median %>% 
  group_by(PTU) %>%
  mutate(max_n = max(n, na.rm = TRUE),  # Calculate max 'n' for each 'PTU'
         normalized_n = n / max_n * 100) %>%  # Normalize 'n' by this max value
  ungroup() %>%  # Remove the grouping
  select(-max_n)  # Optionally remove the max_n column if not needed
# Plot the distribution
p.lines.expectation = ggplot(d.reduced.median.plot, aes(x, median*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome depletion per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.expectation, file='../results/6mer-deviation-expectation.pdf',
       width=10, height=6)
# Plot only first 50kb
ggsave(p.lines.expectation+scale_x_continuous(labels = function(x) x / 1000, limits=c(0,50000)), file='../results/6mer-deviation-expectation-0-50-kb.pdf',
       width=10, height=6)

@@

# And for palindrome density (not expected - actual values)
p.lines.density = ggplot(d.reduced.median.plot, aes(x, median.original*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome density per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.density, file='../results/6mer-palindrome-density.pdf',
       width=10, height=6)

@@

# Order by average score
merge_scores_counts$kmer = ordered(merge_scores_counts$kmer,
                              levels=scores_average$kmer[order(scores_average$score_avg)])
scores.plot = ggplot(merge_scores_counts, aes(score_avg, kmer))+
    geom_vline(xintercept = 0)+
  geom_bar(stat="identity")+
  geom_vline(xintercept = median(merge_scores_counts$score_avg), linetype='dashed', colour='red')+
  theme_bw()+
  theme(axis.text=element_text(colour="black", size=12),
        axis.title.x=element_text(size=24),
        axis.text.x=element_text(size=18))+
  theme(axis.text.y=element_text(hjust=0))+
  ylab("")+
  xlab("Mean depletion score\n(leading vs. lagging, 10kb)")
wilcox.test(merge_scores_counts$score_avg)
ggsave(scores.plot, file='../results/6mer-palindromes-scores.pdf',
       height=10, width=10)
```

>>>>

```{r}
library(ggplot2)
library(knitr)
library(xtable)
library(dplyr)
library(tidyr)
```
### Figure 1

Figure 1. Summary of the 36 plasmid taxonomic units (PTUs) analysed in this paper.

```{r}
# Get average size of PTU
ptu = read.csv('../data/PTU-pangenome-stats.csv')
for (each_ptu in ptu$PTU){
  gene_pa_file <- list.files(paste0("/Users/Liam/Downloads/06_plaspan/", each_ptu), 
                        pattern = "gene_presence_absence.Rtab$", full.names = TRUE, recursive = TRUE)
  gene_counts = read.csv(gene_pa_file, sep='\t', row.names = 1)
  ptu$median.genes[which(ptu$PTU==each_ptu)] = median(colSums(gene_counts))
  ptu$max.genes[which(ptu$PTU==each_ptu)] = max(colSums(gene_counts))
  ptu$min.genes[which(ptu$PTU==each_ptu)] = min(colSums(gene_counts))
}

# Plot these
ptu_for_plotting = ptu %>% pivot_longer(cols=c("n.core", "n.others"))

ptu = ptu[order(ptu$median.genes),]
ptu$PTU = ordered(ptu$PTU, 
                  levels=rev(ptu$PTU))
ptu$leading.region = ifelse(ptu$PTU %in% leading.region.PTUs,
                            "yes", "no")
pdf('../results/Figure-1-ptu-summary-pangenome-plot.pdf', width=9, height=5)
ggplot(ptu[which(ptu$median.genes>20),], aes(PTU, median.genes, ymin=min.genes, ymax=max.genes,
                                             fill=leading.region))+
  geom_bar(aes(y=n.core), stat="identity")+
  geom_errorbar()+
  geom_point(shape=21)+
  theme_bw()+
  xlab("")+
  ylab("Number of genes")+
  labs(fill="Leading region analysed")+
  coord_flip()+
  scale_color_brewer(palette="Set2")+
  scale_fill_brewer(palette="Set2")+
  theme(axis.text=element_text(colour="black"))
dev.off()

@@

```{r}
# Pangenome statistics for PTUs
#ptu = read.csv('../data/PTU-pangenome-stats.csv')

@@

# Maximum counts
d.reduced.median.plot = d.reduced.median %>% 
  group_by(PTU) %>%
  mutate(max_n = max(n, na.rm = TRUE),  # Calculate max 'n' for each 'PTU'
         normalized_n = n / max_n * 100) %>%  # Normalize 'n' by this max value
  ungroup() %>%  # Remove the grouping
  select(-max_n)  # Optionally remove the max_n column if not needed
# Plot the distribution
p.lines.expectation = ggplot(d.reduced.median.plot, aes(x, median*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome depletion per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.expectation, file='../results/Figure-S2-6mer-deviation-expectation.pdf',
       width=10, height=6)
# Plot only first 50kb
ggsave(p.lines.expectation+scale_x_continuous(labels = function(x) x / 1000, limits=c(0,50000)), file='../results/6mer-deviation-expectation-0-50-kb.pdf',
       width=10, height=6)

@@

# And for palindrome density (not expected - actual values)
p.lines.density = ggplot(d.reduced.median.plot, aes(x, median.original*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome density per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.density, file='../results/Figure-S3-6mer-palindrome-density.pdf',
       width=10, height=6)

@@

# Order by average score
merge_scores_counts$kmer = ordered(merge_scores_counts$kmer,
                              levels=scores_average$kmer[order(scores_average$score_avg)])
scores.plot = ggplot(merge_scores_counts, aes(score_avg, kmer))+
    geom_vline(xintercept = 0)+
  geom_bar(stat="identity")+
  geom_vline(xintercept = median(merge_scores_counts$score_avg), linetype='dashed', colour='red')+
  theme_bw()+
  theme(axis.text=element_text(colour="black", size=12),
        axis.title.x=element_text(size=24),
        axis.text.x=element_text(size=18))+
  theme(axis.text.y=element_text(hjust=0))+
  ylab("")+
  xlab("Mean depletion score\n(leading vs. lagging, 10kb)")
wilcox.test(merge_scores_counts$score_avg)
ggsave(scores.plot, file='../results/Figure-S4-6mer-palindromes-scores.pdf',
       height=10, width=10)
```
